---
title: "Chaotic dynamical systems"
author: "Luis Ramón Munive Hernández & Rodrigo Zúñiga Trejo"
date: "9/3/2021"
output: 
  pdf_document:
    keep_tex: true
    latex_engine: xelatex
header-inclues: \usepackage{amsmath, amsthm, amssymb}
---

The Lorenz dynamical system is a system of ordinary differential equations defined as:

\begin{align*}
\frac{\mathrm{d} x}{\mathrm{d} t} & = \sigma (y - x) \\
\frac{\mathrm{d} y}{\mathrm{d} t} & = \rho x - y - x z \\
\frac{\mathrm{d} z}{\mathrm{d} t} & = -\beta z + x y \\
\end{align*}

The default selection for this system parameters is $\sigma = 10, \rho = 28, \beta = 8 / 3$; is known to produce a deterministic chaotic time series.

```{r}
# Needed libraries
library(deSolve)
library(ggplot2)
library(gg3D)
library(latex2exp)
library(nonlinearTseries)
library(TDA)
library(zoo)
library(tseries)
library(dplyr)
library(lubridate)
```

```{r}
# Function for plot path point cloud constructed via Takens' theorem
make_plot <- function(df, theta, phi, title){
  ggplot(df, aes(x = x, y = y, z = z, colour = time)) +
    axes_3D(theta = theta, phi = phi) +
    stat_3D(theta = theta, phi = phi, geom = "path", size = 1) +
    labs_3D(theta = theta, phi = phi,
            labs= c("x", "y", "z"),
            angle = c(0, 0, 0),
            hjust = c(0, 1, 1),
            vjust = c(1, 1, 0), size = 7) +
    ggtitle(title) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
}
```

```{r}
# Lorenz dynamical system

## Parameters
parms <- c(sigma = 10, beta = 8 / 3, rho = 28)
## Initial conditions
state <- c(x = -13, y = -14, z = 47)
## Time span
times <- seq(0, 50, by = 1 / 100)
## Lorenz system differential equations
lorenz <- function(times, state, parms){
  with(data = as.list(c(state, parms)), expr = {
    dxdt <- sigma * (y - x)
    dydt <- rho * x - y - x * z
    dzdt <- x * y - beta * z
    return(list(c(dxdt, dydt, dzdt)))
  })
}

## Making dataframe
lorenz.df <- as.data.frame(ode(y = state, times = times, func = lorenz, parms = parms))
## Lorenz time series for x variable
lorenz.x <- lorenz.df$x
```

```{r Lorenz time series for x, fig.height=3, fig.width=7}
## Ploting Lorenz time series for x variable
plot(lorenz.x, type = "l", lwd = 1, xlab = "Time", ylab = "x", main = "Time series for x variable in Lorenz dynamical system", las = 1)
```

```{r Estimating tau and embedding dimension for Lorenz dynamical system, fig.height=3, fig.width=5}
# Estimating tau and embedding dimension for construct point cloud

## tau-delay estimation based on the mutual information function
tau.lorenz <- timeLag(lorenz.x, technique = "ami", lag.max = 5000, do.plot = T)
tau.lorenz
## embedding dimension computed by using the Cao’s algorithm
#embdim.lorenz <- estimateEmbeddingDim(lorenz.x, time.lag = tau.lorenz, max.embedding.dim = 30)
#embdim.lorenz
## phase space reconstruction via Takens' theorem
tak.lorenz <- buildTakens(lorenz.x, embedding.dim = 3, time.lag = tau.lorenz)
tak.lorenz.df <- as.data.frame(cbind(c(1:dim(tak.lorenz)[1]), tak.lorenz))
colnames(tak.lorenz.df) <- c("time", "x", "y", "z")
```

```{r Path of point cloud for Lorenz time series constructed via Takens theorem, fig.height=6, fig.width=7, warning=FALSE}
make_plot(tak.lorenz.df, theta = 45, phi = 0, title = "Takens' delay embedding attractor reconstruction of the\nstrange chaotic dynamics of Lorenz system") +
  annotate(geom = "text", x = 0.05, y = -0.17, hjust = 0, label = TeX("$\\frac{d x}{d t} = \\sigma (y - x)$"), size = 7) +
  annotate(geom = "text", x = 0.05, y = -0.27, hjust = 0, label = TeX("$\\frac{d y}{d t} = \\rho x - y - x z$"), size = 7) +
  annotate(geom = "text", x = 0.05, y = -0.37, hjust = 0, label = TeX("$\\frac{d z}{d t} = - \\beta z + x y$"), size = 7) +
  theme(plot.title = element_text(size = 15, hjust = 0.5))
```

```{r}
# Lyapunov exponent for Lorenz system
sampling.period.lorenz <- diff(x = lorenz.df$time)[1]
lyapunov.exp.lorenz <- maxLyapunov(time.series = lorenz.x,
                                   sampling.period = sampling.period.lorenz,
                                   min.embedding.dim = 3,
                                   max.embedding.dim = 3 + 3,
                                   time.lag = tau.lorenz,
                                   radius = 1,
                                   max.time.steps = 1000,
                                   do.plot = FALSE)
```

```{r Lyapunov exponent for Lorenz system, fig.width=7, fig.height=5}
plot(lyapunov.exp.lorenz, type = "l", xlim = c(0, 8))
lyapunov.exp.lorenz.est <- estimate(x = lyapunov.exp.lorenz, regression.range = c(0, 3), do.plot = T, type = "l")
lyapunov.exp.lorenz.est
```

```{r}
# Sampling 500 rows of phase space reconstruction, it make cheaper the computation of persistence diagram
set.seed(02092021)
sample.tak.lorenz <- tak.lorenz[sample(x = 1:dim(tak.lorenz)[1], size = 500, replace = FALSE), ]
# persistence diagram construction
#diag.lorenz <- ripsDiag(sample.tak.lorenz[, c(1, 2)], maxdimension = 1, maxscale = max(dist(sample.tak.lorenz[, c(1, 2)])))
diag.lorenz <- ripsDiag(sample.tak.lorenz, maxdimension = 2, maxscale = 7)
```

```{r Persistence diagram for Lorenz system, fig.height=4, fig.width=3.9}
# plot persistence diagram
plot(diag.lorenz$diagram, main = "Persistence diagram of\n Lorenz system")
legend(x = "bottomright", legend = c("0-cycles", "1-cycles", "2-cycles"), lty = c(NA, NA, NA), pch = c(19, 2, 5), col = c("black", "violetred2", "deepskyblue3"), cex = 1, lwd = 2)
```

The Rössler dynamical system is a system of ordinary differential equations defined as:

\begin{align*}
\frac{\mathrm{d} x}{\mathrm{d} t} & = - y - z \\
\frac{\mathrm{d} y}{\mathrm{d} t} & = x + a y \\
\frac{\mathrm{d} z}{\mathrm{d} t} & = b + z (x - c) \\
\end{align*}

The default selection for this system parameters is $a = 0.2, b = 0.2, c = 5.7$; is known to produce a deterministic chaotic time series.

```{r}
# Rössler dynamical system

## Parameters
parms <- c(a = 0.2, b = 0.2, c = 5.7)
## Initial conditions
state <- c(x = -2, y = -10, z = 0.2)
## Time span
times <- seq(0, 50, by = 1 / 100)
## Rössler system differential equations
rossler <- function(times, state, parms){
  with(data = as.list(c(state, parms)), expr = {
    dxdt <- -(y + z)
    dydt <- x + a * y
    dzdt <- b + z * (x - c)
    return(list(c(dxdt, dydt, dzdt)))
  })
}

## Make dataframe
rossler.df <- as.data.frame(ode(y = state, times = times, func = rossler, parms = parms))
## Rössler time series for z variable
rossler.x <- rossler.df$x
```

```{r Rössler time series for x, fig.height=3, fig.width=7}
## Ploting Lorenz time series for x variable
plot(rossler.x, type = "l", lwd = 1, xlab = "Time", ylab = "x", main = "Time series for x variable in Rössler dynamical system", las = 1)
```

```{r Estimating tau and embedding dimension for Rössler dynamical system, fig.height=3, fig.width=5}
# Estimating tau and embedding dimension for construct point cloud

## tau-delay estimation based on the mutual information function
tau.rossler <- timeLag(rossler.x, technique = "ami", lag.max = 5000, do.plot = T)
tau.rossler
## phase space reconstruction via Takens' theorem
tak.rossler <- buildTakens(rossler.x, embedding.dim = 3, time.lag = tau.rossler)
tak.rossler.df <- as.data.frame(cbind(c(1:dim(tak.rossler)[1]), tak.rossler))
colnames(tak.rossler.df) <- c("time", "x", "y", "z")
```

```{r Path of point cloud for Rössler time series constructed via Takens theorem, fig.height=6, fig.width=7, warning=FALSE}
make_plot(tak.rossler.df, theta = 45, phi = 0, title = "Takens' delay embedding attractor reconstruction of the\nstrange chaotic dynamics of Rössler system") +
  annotate(geom = "text", x = 0.05, y = -0.17, hjust = 0, label = TeX("$\\frac{d x}{d t} = - y - z$"), size = 7) +
  annotate(geom = "text", x = 0.05, y = -0.27, hjust = 0, label = TeX("$\\frac{d y}{d t} = x + a y$"), size = 7) +
  annotate(geom = "text", x = 0.05, y = -0.37, hjust = 0, label = TeX("$\\frac{d z}{d t} = b + z (x - c)$"), size = 7) +
  theme(plot.title = element_text(size = 15, hjust = 0.5))
```

```{r}
# Lyapunov exponent for Rössler system
sampling.period.rossler <- diff(x = rossler.df$time)[1]
lyapunov.exp.rossler <- maxLyapunov(time.series = rossler.x,
                                   sampling.period = sampling.period.rossler,
                                   min.embedding.dim = 3,
                                   max.embedding.dim = 3 + 3,
                                   time.lag = tau.rossler,
                                   radius = 1,
                                   max.time.steps = 1000,
                                   do.plot = FALSE)
```

```{r Lyapunov exponent for Rössler system, fig.width=7, fig.height=5}
plot(lyapunov.exp.rossler, type = "l", xlim = c(0, 8))
lyapunov.exp.rossler.est <- estimate(x = lyapunov.exp.rossler, regression.range = c(0, 3), do.plot = T, type = "l")
lyapunov.exp.rossler.est
```

```{r}
# Sampling 500 rows of phase space reconstruction, it make cheaper the computation of persistence diagram
set.seed(02092021)
sample.tak.rossler <- tak.rossler[sample(x = 1:dim(tak.rossler)[1], size = 500, replace = FALSE), ]
# persistence diagram construction
diag.rossler <- ripsDiag(sample.tak.rossler, maxdimension = 2, maxscale = 7)
```

```{r Persistence diagram for Rössler system, fig.height=4, fig.width=3.9}
# plot persistence diagram
plot(diag.rossler$diagram, main = "Persistence diagram of\n Rössler system")
legend(x = "bottomright", legend = c("0-cycles", "1-cycles", "2-cycles"), lty = c(NA, NA, NA), pch = c(19, 2, 5), col = c("black", "violetred2", "deepskyblue3"), cex = 1, lwd = 2)
```

