---
title: "Nonchaotic dynamical systems"
author: "Luis Ramón Munive Hernández & Rodrigo Zúñiga Trejo"
date: "9/3/2021"
output: 
  pdf_document:
    keep_tex: true
    latex_engine: xelatex
header-inclues: \usepackage{amsmath, amsthm, amssymb}
---

The following system belongs to four-term dynamical system with two quadratic term family [Heidel, J., & Fu, Z. (1999)]. It is a system of ordinary differential equations defined as:

\begin{align*}
\frac{\mathrm{d} x}{\mathrm{d} t} & = y^2 + y z\\
\frac{\mathrm{d} y}{\mathrm{d} t} & = x \\
\frac{\mathrm{d} z}{\mathrm{d} z} & = y \\
\end{align*}

The system has a nonchaotic behaviour.

```{r}
# Needed libraries
library(deSolve)
library(ggplot2)
library(gg3D)
library(latex2exp)
library(nonlinearTseries)
library(TDA)
library(zoo)
library(tseries)
library(dplyr)
library(lubridate)
```

```{r}
# Function for plot path point cloud constructed via Takens' theorem
make_plot <- function(df, theta, phi, title){
  ggplot(df, aes(x = x, y = y, z = z, colour = time)) +
    axes_3D(theta = theta, phi = phi) +
    stat_3D(theta = theta, phi = phi, geom = "path", size = 1) +
    labs_3D(theta = theta, phi = phi,
            labs= c("x", "y", "z"),
            angle = c(0, 0, 0),
            hjust = c(0, 1, 1),
            vjust = c(1, 1, 0), size = 7) +
    ggtitle(title) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
}
```

```{r}
# Four-term dynamical system

## Parameters
parms <- c(a = 1, b = 1, c = 1)
## Initial conditions
state <- c(x = 0.01, y = 0, z = 0)
## Time span
times <- seq(0, 10, by = 1 / 100)
## Four-term system differential equations
four.term <- function(times, state, parms){
  with(data = as.list(c(state, parms)), expr = {
    dxdt <- a * y^2 + y * z
    dydt <- b * x
    dzdt <- c * y
    return(list(c(dxdt, dydt, dzdt)))
  })
}

## Make dataframe
four.term.df <- as.data.frame(ode(y = state, times = times, func = four.term, parms = parms))
## Four-term time series for y variable
four.term.y <- four.term.df$y
```

```{r Four-term time series for x, fig.height=3, fig.width=7}
## Ploting Lorenz time series for x variable
plot(four.term.y, type = "l", lwd = 1, xlab = "Time", ylab = "y", main = "Time series for y variable in four-term dynamical system", las = 1)
```

```{r Estimating tau and embedding dimension for four-term dynamical system, fig.height=3, fig.width=5}
# Estimating tau and embedding dimension for construct point cloud

## tau-delay estimation based on the mutual information function
tau.four.term <- timeLag(four.term.y, technique = "ami", lag.max = 5000, do.plot = T)
tau.four.term
## phase space reconstruction via Takens' theorem
tak.four.term <- buildTakens(four.term.y, embedding.dim = 3, time.lag = tau.four.term)
tak.four.term.df <- as.data.frame(cbind(c(1:dim(tak.four.term)[1]), tak.four.term))
colnames(tak.four.term.df) <- c("time", "x", "y", "z")
```

```{r Path of point cloud for four-term time series constructed via Takens theorem, fig.height=6, fig.width=7}
make_plot(tak.four.term.df, theta = 45, phi = 0, title = "Takens' delay embedding attractor reconstruction of the\n nonchaotic dynamics of four-term system") +
  annotate(geom = "text", x = 0.05, y = -0.17, hjust = 0, label = TeX("$\\frac{d x}{d t} = y^2 + y z$"), size = 7) +
  annotate(geom = "text", x = 0.05, y = -0.27, hjust = 0, label = TeX("$\\frac{d y}{d t} = x$"), size = 7) +
  annotate(geom = "text", x = 0.05, y = -0.37, hjust = 0, label = TeX("$\\frac{d z}{d t} = y$"), size = 7) +
  theme(plot.title = element_text(size = 15, hjust = 0.5))
```

```{r}
# Lyapunov exponent for four-term system
sampling.period.four.term <- diff(x = four.term.df$time)[1]
lyapunov.exp.four.term <- maxLyapunov(time.series = four.term.y,
                                   sampling.period = sampling.period.four.term,
                                   min.embedding.dim = 3,
                                   max.embedding.dim = 3 + 3,
                                   time.lag = tau.four.term,
                                   radius = 1,
                                   max.time.steps = 1000,
                                   do.plot = FALSE)
```

```{r Lyapunov exponent for four-term system, fig.width=7, fig.height=5}
plot(lyapunov.exp.four.term, type = "l", xlim = c(0, 8))
lyapunov.exp.four.term.est <- estimate(x = lyapunov.exp.four.term, regression.range = c(0, 3), do.plot = T, type = "l")
lyapunov.exp.four.term.est
```

```{r}
# Sampling 200 rows of phase space reconstruction, it make cheaper the computation of persistence diagram
set.seed(02092021)
sample.tak.four.term <- tak.four.term[sample(x = 1:dim(tak.four.term)[1], size = 200, replace = FALSE), ]
# persistence diagram construction
diag.four.term <- ripsDiag(sample.tak.four.term, maxdimension = 2, maxscale = 3)
```

```{r Persistence diagram for four-term system, fig.height=4, fig.width=3.9}
# plot persistence diagram
plot(diag.four.term$diagram, main = "Persistence diagram for\n four-term dynamical system")
legend(x = "bottomright", legend = c("0-cycles", "1-cycles", "2-cycles"), lty = c(NA, NA, NA), pch = c(19, 2, 5), col = c("black", "violetred2", "deepskyblue3"), cex = 1, lwd = 2)
```

KIC 5520878 star 

```{r}
star <- read.csv("MAST_2021-09-04T2349", header = TRUE)
star <- star[!is.na(star$SAP_FLUX), ]
star.df <- as.data.frame(star)
sap.flux <- star$SAP_FLUX / 10000
```

```{r KIC 5520878 star time series of SAP flux variable, fig.height=3, fig.width=7}
## Ploting KIC 5520878 time series for SAP flux variable
plot(star.df$SAP_FLUX / 10000 , type = "l", lwd = 1, xlab = "Time", ylab = "SAP flux", main = "Time series for Simple Aperture Photometry flux variable in\nKIC 5520878 star dynamics", las = 1)
```

```{r Estimating tau and embedding dimension for KIC 5520878 star time series, fig.height=3, fig.width=5}
# Estimating tau and embedding dimension for construct point cloud

## tau-delay estimation based on the mutual information function
tau.star <- timeLag(sap.flux, technique = "ami", lag.max = 20000, do.plot = T)
tau.star
## phase space reconstruction via Takens' theorem
tak.star <- buildTakens(sap.flux, embedding.dim = 3, time.lag = tau.star)
tak.star.df <- as.data.frame(cbind(c(1:dim(tak.star)[1]), tak.star))
colnames(tak.star.df) <- c("time", "x", "y", "z")
```

```{r Path of point cloud for KIC 5520878 time series constructed via Takens theorem, fig.height=6, fig.width=7}
make_plot(tak.star.df, theta = 45, phi = 0, title = "Takens' delay embedding attractor reconstruction of the\n nonchaotic dynamics of KIC 5520878 star") +
  theme(plot.title = element_text(size = 15, hjust = 0.5))
```

```{r}
# Lyapunov exponent for KIC 5520878 star time series
sampling.period.star <- diff(x = tak.star.df$time)[1]
lyapunov.exp.star <- maxLyapunov(time.series = sap.flux,
                                   sampling.period = sampling.period.star,
                                   min.embedding.dim = 3,
                                   max.embedding.dim = 3 + 3,
                                   time.lag = tau.star,
                                   radius = 1,
                                   max.time.steps = 1000,
                                   do.plot = FALSE)
```

```{r Lyapunov exponent for KIC 5520878 star time series, fig.width=7, fig.height=5}
plot(lyapunov.exp.star, type = "l", xlim = c(0, 8))
lyapunov.exp.star.est <- estimate(x = lyapunov.exp.star, regression.range = c(0, 3), do.plot = T, type = "l")
lyapunov.exp.star.est
```

```{r}
# Sampling 200 rows of phase space reconstruction, it make cheaper the computation of persistence diagram
set.seed(02092021)
sample.tak.star <- tak.star[sample(x = 1:dim(tak.star)[1], size = 200, replace = FALSE), ]
# persistence diagram construction
diag.star <- ripsDiag(sample.tak.star, maxdimension = 2, maxscale = 1)
```

```{r Persistence diagram for KIC 5520878 star time series, fig.height=4, fig.width=3.9}
# plot persistence diagram
plot(diag.star$diagram, main = "Persistence diagram for\n KIC 5520878 star time series")
legend(x = "bottomright", legend = c("0-cycles", "1-cycles", "2-cycles"), lty = c(NA, NA, NA), pch = c(19, 2, 5), col = c("black", "violetred2", "deepskyblue3"), cex = 1, lwd = 2)
```

References

Heidel, J., & Fu, Z. (1999). Nonchaotic behaviour in three-dimensional quadratic systems II. The conservative case. Nonlinearity, 12(3), 617.